---
title: "Plotting Omics Data"
author: "Jiratchaya Nuanpirom"
date: "17 May 2023"
format:
  html:
    toc: true
    number-sections: true
    number-depth: 3
execute: 
  eval: true
  echo: true
  warning: false
---

Load R pakcages

```{r}
library(tidyverse)
library(paletteer)
library(scales)
library(RColorBrewer)
library(ggpubr)
library(VennDiagram)
library(circlize)
library(viridis)
library(png)
library(GGally)
library(cowplot)
library(openxlsx)
library(gridExtra)
library(gtable)
library(egg)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(gridGraphics)
library(corrplot)
library(plotly)
library(agricolae)
library(ComplexHeatmap)
library(DESeq2)
library(pcaExplorer)
library(EnhancedVolcano)
```

## Plotting Abundance of Features

### Abundance of Representative Terms

### Relative abundance

## Principal Component Analysis

Way 1: prcomp & normal scatter plot

Way 2: rlogtransform & pcaexplorer

## Hierarchical Clustering Analysis (Heatmap)

```{r}
set.seed(123)
nr1 = 4; nr2 = 8; nr3 = 6; nr = nr1 + nr2 + nr3
nc1 = 6; nc2 = 8; nc3 = 10; nc = nc1 + nc2 + nc3
mat = cbind(rbind(matrix(rnorm(nr1*nc1, mean = 1,   sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc1, mean = 0,   sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc1, mean = 0,   sd = 0.5), nr = nr3)),
    rbind(matrix(rnorm(nr1*nc2, mean = 0,   sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc2, mean = 1,   sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc2, mean = 0,   sd = 0.5), nr = nr3)),
    rbind(matrix(rnorm(nr1*nc3, mean = 0.5, sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc3, mean = 0.5, sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc3, mean = 1,   sd = 0.5), nr = nr3))
   )
mat = mat[sample(nr, nr), sample(nc, nc)] # random shuffle rows and columns
rownames(mat) = paste0("row", seq_len(nr))
colnames(mat) = paste0("column", seq_len(nc))

library(circlize)
col_fun = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
col_fun(seq(-3, 3))

Heatmap(mat, name = "mat", col = col_fun)
```

## Volcano Plot

Enhancedvolcano

## Venn Diagram and Upset Plot

Venn: venndiagram

```{r}

```

upset: complexheatmap

```{r}
library(ComplexHeatmap)
set.seed(123)
lt = list(a = sample(letters, 5),
          b = sample(letters, 10),
          c = sample(letters, 15))
m1 = make_comb_mat(lt)
UpSet(m1)
```

## Plotting Diversity Indices

ggpubr

## Plotting Individual Gene Expression

ggpubr

## Circos plot

## 

## Session info

```{r}
sessionInfo()
```
